manuscript_figures

### 2024-06-28 scripts/data_cleaning.Rmd
updated nmds script and plot

# Combine QUAD dataframes
```{r}

#make QUAD data long
# Function to rename columns using metadata
# Get the unique column names from the QUAD dataframe
column_names.quad <- colnames(QUAD)
# Create a lookup table with name-to-code mappings
name_to_code.quad <- setNames(species.metadata$code, species.metadata$name)

>>>>

# Combine QUAD dataframes
```{r}
#make QUAD data long
# Function to rename columns using metadata
# Get the unique column names from the QUAD dataframe
column_names.quad <- colnames(QUAD)
# Create a lookup table with name-to-code mappings
name_to_code.quad <- setNames(species.metadata$code, species.metadata$name)

### 2024-06-28 scripts/manuscript_figures.Rmd
updated nmds script and plot

# remove control column and two last columns, not needed
quad.filt <- quad.filt[ -c(13:14) ] 
quad.filt <- quad.filt[ -c(8) ] 
quad.filt <- quad.filt[ -c(5:6) ]

@@

# Make line plot for temperature time series by site per month
```{r}
# Calculate mean and standard error of temperature by month and site
avg_temp <- temperature_data %>%
  mutate(MonthAbbrev = month.abb[Month]) %>%  # Add month abbreviations
  group_by(MonthAbbrev, Site) %>%
  summarize(
    avg_temp = mean(Temp),
    std_error = sd(Temp) / sqrt(n()))  # Calculate standard error

@@

# Plotting average temperature by site as lines and points per month with error bars
temperature_plot <- ggplot(avg_temp, aes(x = MonthAbbrev, y = avg_temp, color = Site, group = Site)) +
  geom_line(size = 1.5) +  # Lines for average temperature by site
  geom_point(size = 3) +  # Points for mean values by site
  geom_errorbar(aes(ymin = avg_temp - std_error, ymax = avg_temp + std_error), width = 0.2) +  # Error bars for SEM
  labs(x = "Month", y = expression("Average Temperature (°C)"), color = "Site") +
  scale_color_manual(values = c("Fort Wetherill" = "grey", "King's Beach" = "black")) +  # Custom colors
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18));temperature_plot

# Save the plot to a file
ggsave("output/figures/temp.plot.png", plot = temperature_plot, width = 20, height = 10)
```

@@

# Ensure abundance columns are numeric
wide_data <- wide_data %>%
  mutate(across(crustacean:molluscs, as.numeric))

# Calculate Bray-Curtis dissimilarity
m_com <- vegdist(wide_data[, -c(4:6)], method = "bray")

@@

# Create a dataframe for plotting
plot_data <- data.frame(
  NMDS1 = nmds$points[, 1],  # NMDS dimension 1
  NMDS2 = nmds$points[, 2],  # NMDS dimension 2
  type = invert.count$type,  # Type of invertebrate
  abundance = invert.count$abundance  # Abundance of invertebrate
)

# Plot nMDS with type labels and scaled points by abundance
ggplot(plot_data, aes(x = NMDS1, y = NMDS2, color = type, size = abundance)) +
  geom_point(alpha = 0.8) +
  scale_size_continuous(range = c(2, 8)) +  # Adjust point size range as needed
  labs(title = "nMDS Plot of Invertebrate Types by Abundance",
       x = "NMDS 1", y = "NMDS 2") +
  theme_minimal() +
  theme(legend.position = "right") +
  geom_text(aes(label = type), nudge_x = 0.05, nudge_y = 0.05, size = 3)  # Add type labels
```

>>>>

# remove control column and two last columns, not needed
quad.filt <- quad.filt[ -c(9) ] 
#quad.filt <- quad.filt[ -c(8) ] 
#quad.filt <- quad.filt[ -c(5:6) ]

@@

# Make line plot for temperature time series by site per year
```{r}
# Calculate mean and standard error of temperature by month, site, and year
avg_temp <- temperature_data %>%
  mutate(MonthAbbrev = month.abb[Month]) %>%  # Add month abbreviations
  group_by(Year, MonthAbbrev, Site) %>%
  summarize(
    avg_temp = mean(Temp),
    std_error = sd(Temp) / sqrt(n()),  # Calculate standard error
    .groups = 'drop'  # Drop grouping after summarization
  )

@@

# Ensure MonthAbbrev is a factor with levels in the correct chronological order
avg_temp$MonthAbbrev <- factor(avg_temp$MonthAbbrev, levels = month.abb)

# Plotting average temperature by site as lines and points per month with error bars
temperature_plot <- ggplot(avg_temp, aes(x = MonthAbbrev, y = avg_temp, color = Site, group = interaction(Site, Year))) +
  geom_line(size = 1.5) +  # Lines for average temperature by site and year
  geom_point(size = 3) +  # Points for mean values by site and year
  geom_errorbar(aes(ymin = avg_temp - std_error, ymax = avg_temp + std_error), width = 0.2) +  # Error bars for SEM
  labs(x = "Month", y = expression("Average Temperature (°C)"), color = "Site") +
  facet_wrap(~ Year) +  # Facet by year
  scale_color_manual(values = c("Fort Wetherill" = "grey", "King's Beach" = "black")) +  # Custom colors
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18));temperature_plot

# Save the plot to a file
ggsave("output/figures/temp.plot.year.png", plot = temperature_plot, width = 20, height = 10)
```

# Make line plot for temperature time series by site per month
```{r}
# Calculate mean and standard error of temperature by month, site, and year
avg_temp_month <- temperature_data %>%
  mutate(MonthAbbrev = month.abb[Month]) %>%  # Add month abbreviations
  group_by(MonthAbbrev, Site) %>%
  summarize(
    avg_temp = mean(Temp),
    std_error = sd(Temp) / sqrt(n()))  # Calculate standard error  # Drop grouping after summarization

# Map site codes to full names
avg_temp_month$Site <- ifelse(avg_temp_month$Site == "FW", "Fort Wetherill", "King's Beach")

# Ensure MonthAbbrev is a factor with levels in the correct chronological order
avg_temp_month$MonthAbbrev <- factor(avg_temp_month$MonthAbbrev, levels = month.abb)

# Plotting average temperature by site as lines and points per month with error bars
temperature_plot_month <- ggplot(avg_temp_month, aes(x = MonthAbbrev, y = avg_temp, color = Site, group = Site)) +
  geom_line(size = 1.5) +  # Lines for average temperature by site
  geom_point(size = 3) +  # Points for mean values by site
  geom_errorbar(aes(ymin = avg_temp - std_error, ymax = avg_temp + std_error), width = 0.2) +  # Error bars for SEM
  labs(x = "Month", y = expression("Average Temperature (°C)"), color = "Site") +
  scale_color_manual(values = c("Fort Wetherill" = "grey", "King's Beach" = "black")) +  # Custom colors
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18));temperature_plot_month

# Save the plot to a file
ggsave("output/figures/temp.plot.month.png", plot = temperature_plot_month, width = 20, height = 10)
```

@@

# Ensure only numeric columns are selected for vegdist
numeric_data <- wide_data[, c("crustacean", "echinoderms", "molluscs")]

# Remove rows where all abundance values are zero
numeric_data <- numeric_data[rowSums(numeric_data) > 0, ]

# Calculate Bray-Curtis dissimilarity
m_com <- vegdist(numeric_data, method = "bray")

@@

# Extract scores
scores <- as.data.frame(scores(nmds))

# Add row numbers as a unique identifier
scores$ID <- row.names(scores)

# Create a data frame with SITE and YEAR information
site_year_info <- wide_data[rowSums(wide_data[, c("crustacean", "echinoderms", "molluscs")]) > 0, c("SITE", "YEAR")]
site_year_info$ID <- row.names(site_year_info)

# Merge SITE and YEAR information into scores
scores <- merge(scores, site_year_info, by = "ID")

# Plotting nMDS results
nmds_plot <- ggplot(scores, aes(x = NMDS1, y = NMDS2, color = SITE, shape = as.factor(YEAR))) +
  geom_point(size = 8) +
  labs(x = "NMDS1", y = "NMDS2", color = "Site", shape = "Year") +
  stat_ellipse(aes(group = YEAR), level = 0.95, linetype = 2, size = 1, alpha = 0.2)+
  scale_color_manual(values = c("Fort Wetherill" = "grey", "King's Beach" = "black")) +
  theme_bw() +
  theme(
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    axis.text.x = element_text(size = 18),
    axis.text.y = element_text(size = 18),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18));nmds_plot
